Water discharge has a variable impact on sediment transport capacity in subglacial systems

### 2023-08-25 Delaneyetal2023b.tex
Updates from Overleaf

    ianarburua.delaney@unil.ch},
  Daniel Farinotti\footnote{Laboratory of Hydraulics, Hydrology and Glaciology (VAW), ETH-Z\"urich, H\"onggerbergring 26, 8093 89 Z\"urich, Switzerland}$^{,}$\footnote{Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) Z\"uricherstrasse 111, 8903 1011 Birmensdorf, Switzerland},
  Andrew J. Tedstone\footnote{Department of Geosciences, University of Fribourg, Ch. du Musée 1700, Fribourg, Switzerland}

@@

Changes in glacier dynamics, geomorphology, and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \citep[e.g.][]{zhang2022}.
Increases in sediment transport  have been observed in Greenland \citep{bendixen2017}, the European Alps \citep{costa2017}, the Himalayas \citep{li2021}, and the Andes \citep{vergara2022}.
In some of these regions, increased water discharge and glacier melt have been interpreted to yeild greater sediment transport capacity \citep{bendixen2017,costa2017,li2021}.
Observed changes to sediment transport in glacierized catchments require examining the processes controlling sediment discharge in these catchments and its variations with water discharge  \citep[e.g.][]{riihimaki2005,swift2005}.

@@

\ian{comeback to Andrew's comments here}
Water and suspended sediment data are leveraged from two sites in this study.
Water discharge and suspended sediment concentration were collected at a $1$\unit{min} interval from the Fieschergletscher in the Swiss Alps  continuously over the period 2014--2021 ($46^\circ\,29'\,07"$ N, $8^\circ\,08'\,34"$ E).
A description of the data and its collection is available in \citet{felix2022} and \citet{felix2021}.

Additionally, water discharge and suspended sediment concentration were collected roughly $2$\unit{km} downstream from Leverett Glacier in Greenland over  periods from June to September from 2009 to 2012 ($67^\circ\,03'\,50"$ N, $50^\circ\,12'\,59"$ W).
This data was presented in part in \citet{cowton2012} and is available from \citet{tedstone2017} at a $5$\unit{min} time interval.
Note that in some cases the removal and relocation of sensors created variations in the sediment concentration not resulting from natural processes.

@@

Over time scales longer than days to weeks, the glacier's sediment transport capacity reacts to the ice thickness, controlling the channel closure rate, and the surface slope of the glacier \citep[Figure~\ref{fig:multi_run}, ~ Section~\ref{sect:sub_mode}; ][]{rothlisberger1972,shreve1972,stevens2022}.
If the glacier is in a supply-limited state, often assumed in landscape evolution models, then the sediment discharge record also represents additional processes of sediment availability and bedrock erosion from  water pressure variations and sliding  \citep[e.g.][]{iverson2012,herman2015,delaney2023}.
This multitude of processes lies in contrast to most subaerial systems, where transport capacity typically increases with water discharge and hydraulic gradient, and the hydraulic gradient generally remains stable over time scales ranging from years to millennia \citep[Section~\ref{sect:fluv}; e.g.][]{muller1968,whipple1999,wickert2019}.
The stronger link between sediment export and hydrological forcing in subaerial channels means that variations in sediment transport records can be more easily attributed to climatic conditions than in  subglacial systems

@@

Implications of the findings here include:
\begin{enumerate}
\item the inconsistent relationship between water discharge and sediment transport capacity underneath glaciers, causing increased variability in sediment transport capacity compared to river systems;
\item the timing of peak sediment transport capacity below glaciers generally coincides with large rates of increasing in water discharge, as opposed to the maximum quantity of water discharge;
\item the possibility for a consistent relationship between water and sediment discharge to represent subglacial sediment access or open channel hydraulics near the glacier margin, instead of subglacial sediment transport capacity.

>>>>

    ianarburua.delaney@unil.ch},
  Andrew J. Tedstone\footnote{Department of Geosciences, University of Fribourg, Ch. du Musée 1700, Fribourg, Switzerland},
  Daniel Farinotti\footnote{Laboratory of Hydraulics, Hydrology and Glaciology (VAW), ETH-Z\"urich, H\"onggerbergring 26, 8093 89 Z\"urich, Switzerland}$^{,}$\footnote{Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) Z\"uricherstrasse 111, 8903 1011 Birmensdorf, Switzerland}

@@

Changes in glacier dynamics, geomorphology, and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \citep[e.g.][]{zhang2022}.
Increases in sediment transport  have been observed in Greenland \citep{bendixen2017}, the European Alps \citep{costa2017}, the Himalayas \citep{li2021}, and the Andes \citep{vergara2022}.
In some of these regions, increased water discharge and glacier melt have been interpreted to yield greater sediment transport capacity \citep{bendixen2017,costa2017,li2021}.
Observed changes to sediment transport in glacierized catchments require examining the processes controlling sediment discharge in these catchments and its variations with water discharge  \citep[e.g.][]{riihimaki2005,swift2005}.

@@

Water and suspended sediment data are leveraged from an Alpine and an ice sheet setting.
The Alpine site is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07"$ N, $8^\circ\,08'\,34"$ E).
Water discharge and suspended sediment concentration were collected here at a $1$\unit{min} interval continuously over the period 2014--2021. 
A description of the data and its collection is available in \citet{felix2022} and \citet{felix2021}.

The Leverett Glacier in Greenland served as the ice sheet setting.
Water discharge and suspended sediment concentration were collected roughly $2$\unit{km} downstream from the terminus over  periods from June to September from 2009 to 2012 ($67^\circ\,03'\,50"$ N, $50^\circ\,12'\,59"$ W).
This data was presented in part in \citet{cowton2012} and is available from \citet{tedstone2017} at a $5$\unit{min} time interval.
Note that in some cases the removal and relocation of sensors created variations in the sediment concentration not resulting from natural processes.

@@

Over time scales longer than days to weeks, the glacier's sediment transport capacity reacts to the ice thickness, controlling the channel closure rate, and the surface slope of the glacier, in addition to water discharge \citep[Figure~\ref{fig:multi_run}, ~ Section~\ref{sect:sub_mode}; ][]{rothlisberger1972,shreve1972,stevens2022}.
If the glacier is in a supply-limited state, often assumed in landscape evolution models, then the sediment discharge record also represents additional processes of sediment availability and bedrock erosion from  water pressure variations and sliding  \citep[e.g.][]{iverson2012,herman2015,delaney2023}.
This multitude of processes lies in contrast to most subaerial systems, where transport capacity typically increases with water discharge and hydraulic gradient, and the hydraulic gradient generally remains stable over time scales ranging from years to millennia \citep[Section~\ref{sect:fluv}; e.g.][]{muller1968,whipple1999,wickert2019}.
The stronger link between sediment export and hydrological forcing in subaerial channels means that variations in sediment transport records can be more easily attributed to climatic conditions than in  subglacial systems

@@

Implications of the findings here include:
\begin{enumerate}
\item the inconsistent relationship between water discharge and sediment transport capacity underneath glaciers, causing increased variability in sediment transport capacity compared to river systems;
\item the timing of peak sediment transport capacity below glaciers generally coincides with large rates of increasing in water discharge, as opposed to the maximum quantity of water discharge;
\item a potentially consistent relationship between water and sediment discharge may represent subglacial sediment access or open channel hydraulics near the glacier margin, instead of subglacial sediment transport capacity.
